---
title: 'EDS 223: assignment 2'
author: "Rosemary Juarez"
output: html_document
date: "2023-10-17"
editor_options: 
  markdown: 
    wrap: 72
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930's the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York
Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.[^4] Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Check out coverage by
[EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

### Data

#### EJScreen

We will be working with data from the United States Environmental
Protection Agency's EJScreen: Environmental Justice Screening and
Mapping Tool.

According to the [US EPA
website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents
> or other stakeholders as they search for environmental or demographic
> information. It can also support a wide range of research and policy
> goals. The public has used EJScreen in many different locations and in
> many different ways.
>
> EPA is sharing EJScreen with the public:\
> - to be more transparent about how we consider environmental justice
> in our work,\
> - to assist our stakeholders in making informed decisions about
> pursuing environmental justice and,\
> - to create a common starting point between the agency and the public
> when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels.
You will be working with block group data that has been downloaded from
the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). To
understand the associated data columns, you will need to explore the
Technical Documentation and column description spreadsheet available in
the data folder. I also encourage you to explore the [limitations and
caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen)
of the data.

#### Mapping Inequality

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

We will be working with maps of HOLC grade designations for Los Angeles.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

#### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed.

We will be working observations of birds from 2021 onward.

## Assignment

### Investigate the legacy of redlining in current environmental (in)justice

Load relevant packages.

```{r }
library(sf)
library(terra)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)
library(tmap)    # for static and interactive maps
library(leaflet) # for interactive maps
library(ggplot2) # tidyverse data visualization package
library(here)
```

Read in EJScreen data and filter to Los Angeles County (5 points)\

```{r}
#
# DO NOT RUN DO NOT RUN DO NOT RUN DO NOT RUN DO NOT RUN DO NOT RUN DO NOT RUN DO NOT 

#note to self: DO NOT RUN OR IT WILL TAKE UP YOUR TIME
#reading ejscreen here
ejscreen <- st_read("C:/Users/rosem//Documents/MEDS/Courses/EDS-223/assignment-2-rosemaryjuarez/data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb")

# DO NOT RUN DO NOT RUN DO NOT RUN DO NOT RUN DO NOT RUN DO NOT RUN DO NOT RUN DO NOT 
```

```{r}

#filtering to los angeles only
los_angeles <- filter(ejscreen, CNTY_NAME == "Los Angeles County" )

#now just viewing it\
View(los_angeles)

#it worked!
```

Make a map of wastewater discharge by census block groups. Indicate
which census block groups are above the 95th percentile of national
values for wastewater discharge by adding a centroid. (10 points)

```{r }
#gotta filter first:
water_95 <- filter(los_angeles, P_PWDIS >95) %>% 
    st_centroid()

#fixing to make it zoom in closer w out islands
los_angeles <- los_angeles %>% 
  filter(P_PWDIS != 'na')

tm_shape(los_angeles) +
  tm_polygons(fill = 'P_PWDIS',
              title = 'Percentile of wastewater discharge') +
  tm_shape(water_95) +
  tm_dots('P_PWDIS',
          fill = 'red') +
  tm_scale_bar() +
  tm_compass() +
  tm_title('Los Angeles Census block: above 95th percentile in Wastewater Discharge')

#yay first map down!

```

Find the percent of census block groups that have:\
- less than 5% of the population is considered low income (5 points)

```{r }

#first setting low_income
low_income <- los_angeles %>% 
  filter(LOWINCPCT < .05)

#calculating the percentage of census blokc groups
(nrow(low_income)/nrow(los_angeles)) * 100
```

Find the percent of census block groups that are:\
- above the 80th percentile for Particulate Matter 2.5 AND\
- above the 80th percentile for Superfund proximity (10 points)

```{r }
#found 80th percentile and above in the same line
pm2_5 <- los_angeles %>% 
  filter((P_PM25 > 80) & (P_PNPL > 80)) %>% 
  View()
```

Import redlining information for Los Angeles.

```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json") %>%
  st_make_valid()

```

Make a map of historical redlining boundaries, colored by HOLC grade. (5
points)

```{r }
#first i want to add a new column for a description of the boundaries

#first need to read in data and see what it looks like
tm_shape(LA_redlining) +
  tm_polygons(fill = 'fill') +
  tm_compass() +
  tm_scale_bar() +
  tm_graticules() +
  tm_title('HOLC Grade in Los Angeles')

```

Find the number of census block groups that fall within areas with HOLC
grades hint: make sure the CRS match (15 points)

```{r eval=FALSE}
#check to see if they match
st_crs(los_angeles) == st_crs(LA_redlining)

#change CRS
LA_redlining <- st_transform(LA_redlining, "EPSG:3857")

#find census block groups using join

holc_census <- st_join(x = los_angeles, y = LA_redlining)

#checking
View(holc_census)
#it seems like it successfully joined

#now going to check how many were successful interections using holc_grades

nrow(holc_census)

#making sure this is right by checking  original data
nrow(los_angeles)
nrow(LA_redlining)
```

Summarize current conditions based on EJScreen data within historical
redlining categories using the mean of the following variables:\
-% low income.\
- percentile for particulate Matter 2.5.\
- percentile for low life expectancy.\
- percentile for air toxics cancer risk (20 points)

```{r }
holc_census %>% 
  group_by(fill) %>% 
  summarise(lowinc_mean = mean(LOWINCPCT, na.rm = TRUE) * 100,
            pm_25_mean = mean(P_PM25, na.rm = TRUE),
            life_exp_mean = mean(P_LIFEEXPPCT, na.rm = TRUE),
            cancer_mean = mean(P_CANCER, na.rm = TRUE)
            )
```

Please write 2-3 sentences reflecting on these results. (5 points)

From the evidence above, Class D consistently has the worst ratings
throughout the table. It not only has the lowest average on
Socio-Economic Status, but it also suffers from higher risk of cancer or
lower life expectancy. Meanwhile class A has consistent top marks in
health and wealth.

### Investigate the legacy of redlining in biodiversity observations

For bird observations from 2022 that fall within neighborhoods with HOLC
grads, find the percent of observations within each redlining categories
and plot results. hint: make sure that the bird observations have the
same CRS as redlining data. (20 points)

```{r}
#forgot to read in data:
birds <- st_read("C:/Users/rosem/Documents/MEDS/Courses/EDS-223/assignment-2-rosemaryjuarez/data/gbif-birds-LA")

#checking CRS again
st_crs(holc_census) == st_crs(birds)

#change CRS
birds <- st_transform(birds, "EPSG:3857")

bird_holc <- st_join(x = holc_census, y = birds)

tm_shape(bird_holc) +
  tm_polygons()
```

Please write 2-3 sentences reflecting on these results. Are they
surprising? If so, why? What are some potential implications? (5 points)
